---
title: "Human Activity Recognition - Weight Lifting Exploratory Data Analysis"
author: "Maurício Collaça"
date: "December, 5th 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Read more: http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises

## Data validation and transformation
```{r message=FALSE, warning=FALSE, include=FALSE}
library(dplyr); library(ggplot2); library(GGally); library(RColorBrewer)
training <- read.csv("pml-training.csv", stringsAsFactors = FALSE)
testing <- read.csv("pml-testing.csv", stringsAsFactors = FALSE)
```

The training set row and columns dimensions
```{r echo=FALSE}
dim(training)
```

The testing set row and columns dimensions
```{r echo=FALSE}
dim(testing)
```

### Missing values: NA, "#DIV/0!" or ""

100 training variables have either 98% or 100% missing values.
```{r echo=FALSE, comment=""}
missProp <- function(x) mean(is.na(x) | (is.character(x) & x %in% c("", "#DIV/0!")))
trainNAproportion <- round(sapply(training, missProp),2)
table(trainNAproportion)
trainBadVars <- names(trainNAproportion[trainNAproportion > 0])
```

100 testing variables have 100% missing values.
```{r echo=FALSE, comment=""}
testNAproportion <- round(sapply(testing, missProp),2)
table(testNAproportion)
testBadVars <- names(testNAproportion[testNAproportion > 0])
```

By visually inspecting the dataset it was concluded the high missing value proportion is due to statistics calculated in time windows and only stored at rows whose new_window = "yes", characterizing a messy data set where rows are not allways representing a single observation.  As those statistics can be recalculated from the respective raw data when and if necessary, they will be ignored.
```{r include=FALSE}
badVars <- unique(c(names(trainNAproportion[trainNAproportion > 0]),
                    names(testNAproportion[testNAproportion > 0])))
training <- select(training, -one_of(badVars))
testing <- select(testing, -one_of(badVars))
training$new_window <- NULL
training$num_window <- NULL
training$cvtd_timestamp <- NULL
testing$new_window <- NULL
testing$num_window <- NULL
testing$cvtd_timestamp <- NULL
classText <- c("A:correctly", "B:throwing elbows", "C:lifting halfway", "D:lowering halfway", "E:throwing hips")
training <- training %>% mutate(classe = factor(classe, labels = classText), user_name = factor(user_name))
```

## Exploratory Data Analysis

### Contingency table

Sensor readings by classification and user
```{r, comment=""}
with(training, rbind(cbind(table(classe, user_name), "(total)" = table(classe)),
                     "(total)" = c(table(user_name), length(user_name))))
```

### Heatmaps
```{r include=FALSE}
blue.pal <- function(n) { f <- colorRampPalette(brewer.pal(9, "Blues")) ; rev(f(n)) }
grey.pal <- function(n) { f <- colorRampPalette(brewer.pal(9, "Greys")) ; rev(f(n)) }
heat.pal <- function(n) { f <- colorRampPalette(brewer.pal(9, "YlOrRd")) ; rev(f(n)) }
```
```{r fig.height=5, fig.width=8, include=FALSE}
imagemap <- function(x, cluster=TRUE, col, main) {
    if(cluster) {
        distance <- dist(x)
        distanceT <- dist(t(x))
        rowInd <- order.dendrogram(reorder(as.dendrogram(hclust(distance)), rowMeans(x)))
        colInd <- order.dendrogram(reorder(as.dendrogram(hclust(distanceT)), colMeans(x)))
        image(t(scale(x[rowInd, colInd])), col=col, axes=FALSE, main=main)
    }
    else {
        image(t(scale(x)), col=col, axes=FALSE, main=main)
    }
    axis(1, at=seq(0,1, length=ncol(x)), labels = colnames(x), las=2, cex.axis=0.7, hadj=0.9, padj=0)
}
```

```{r classUserClusters, echo=FALSE, fig.height=6, fig.width=10}
dataColors <- heat.pal(4)
par(mfrow=c(1,2), mar=c(6,1,2,1))
m <- as.matrix(training[order(training$classe, training$user_name, training$raw_timestamp_part_1), -c(1:4,57)])
imagemap(m, cluster=FALSE, col = dataColors, main="Rows ordered by class, user and timestamp")
abline(h=cumsum(table(training$classe))/sum(table(training$classe)), lty=2, lwd=1)
m <- as.matrix(training[order(training$user_name, training$classe, training$raw_timestamp_part_1), -c(1:4,57)])
imagemap(m, cluster=FALSE, col = dataColors, main="Rows ordered by user, class and timestamp")
abline(h=cumsum(table(training$user_name))/sum(table(training$user_name)), lty=2, lwd=1)
```

Heatmap ordered by hierarchical cluster dendrograms of the Euclidean distances and complete-linkage agglomeration method.
```{r overallCluster, echo=FALSE, fig.height=12, fig.width=12, cache=TRUE}
rowColors <- grey.pal(nrow(m)); colColors <- grey.pal(ncol(m))
heatmap(m, scale="column", margins=c(10,1), cexCol=1, RowSideColors=rowColors, ColSideColors=colColors,
        col=dataColors, labRow=NA, main="Row ordered by ")
```

Heatmaps segmented by sensor set, ordered by hierarchical cluster of Euclidean distances and complete-linkage agglomeration method.
```{r sensorClusters, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
par(mfrow=c(2,2), mar=c(6,1,2,1))
dataColors <- heat.pal(10)
training %>% select(contains("_arm")) %>% as.matrix %>% imagemap(T, dataColors, "Arm")
training %>% select(contains("_belt")) %>% as.matrix %>% imagemap(T, dataColors, "Belt")
training %>% select(contains("_dumbbell")) %>% as.matrix %>% imagemap(T, dataColors, "Dumbbell")
training %>% select(contains("_forearm")) %>% as.matrix %>% imagemap(T, dataColors, "Forearm")
```

### Generalized pairs plots

##### Accelerometer totals by sensor set
```{r totalPairs, echo=FALSE, fig.height=4, fig.width=4, cache=TRUE}
training %>% select(matches("total")) %>% ggpairs(columnLabels = gsub("_", "\n", names(.)))
```

#### Euler angles (pitch, roll and yaw) from sensors (belt, arm, dumbell and Forearm)
```{r anglePairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(matches("^(accel|gyros|magnet|pitch|roll|yaw).+[^xyz]$")) %>% ggpairs(columnLabels = gsub("_", "\n", names(.)))
```

#### Arm sensor set
```{r armPairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(contains("_arm")) %>% ggpairs(columnLabels = gsub("_", "\n", names(.)),
                                                  title="Arm sensor set features")
```
```{r eval=FALSE, comment="", include=FALSE}
training %>% select(ends_with("_arm")) %>% names
training %>% select(matches("_arm_[xyz]$")) %>% names
```

#### Belt sensor set
```{r beltPairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(contains("_belt")) %>% ggpairs(columnLabels = gsub("_", "\n", names(.)),
                                                   title="Belt sensor set features")
```
```{r eval=FALSE, comment="", include=FALSE}
training %>% select(ends_with("_belt"), ends_with("_belt.1")) %>% names
training %>% select(matches("_belt_[xyz]$")) %>% names
```

#### Dumbbell sensor set
```{r dumbbellPairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(contains("_dumbbell")) %>% ggpairs(columnLabels = gsub("_", "\n", names(.)),
                                                       title="Dumbbell sensor set features")
```
```{r eval=FALSE, comment="", include=FALSE}
training %>% select(ends_with("_dumbbell")) %>% names
training %>% select(matches("_dumbbell_[xyz]$")) %>% names
```

#### Forearm sensor set
```{r forearmPairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(contains("_forearm")) %>% ggpairs(columnLabels = gsub("_", "\n", names(.)),
                                                      title="Forearm sensor set features")
```
```{r eval=FALSE, comment="", include=FALSE}
training %>% select(ends_with("_forearm")) %>% names
training %>% select(matches("_forearm_[xyz]$")) %>% names
```

#### Axial signals

##### Arm axial signals
```{r axialArmPairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(matches("_arm_[xyz]$")) %>%
    ggpairs(columnLabels = gsub("_", "\n", names(.)),
            title="Arm, Belt, Dumbbell and Forearm sensor set axial (XYZ) features")
```

##### Belt axial signals
```{r axialBeltPairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(matches("_belt_[xyz]$")) %>%
    ggpairs(columnLabels = gsub("_", "\n", names(.)))
```

##### Dumbbell axial signals
```{r axialDumbbellPairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(matches("_dumbbell_[xyz]$")) %>%
    ggpairs(columnLabels = gsub("_", "\n", names(.)))
```

##### Forearm axial signals
```{r axialForearmPairs, echo=FALSE, fig.height=9, fig.width=9, cache=TRUE}
training %>% select(matches("_forearm_[xyz]$")) %>%
    ggpairs(columnLabels = gsub("_", "\n", names(.)))
```



